The bhsdtr package: a general purpose method of Bayesian inference for Signal Detection Theory models

### 2019-03-13 inst/preprint/paper.tex
kosmetyczne zmiany w tekscie

However, because the SDT model is non-linear, variability in its
parameters due to factors such as participants or items has to be
accounted for, otherwise the estimates of SDT parameters are
biased. As we explain later in this paper, neither of the two
available methods of inference for hierarchical SDT models that we are
aware of addresses this problem correctly. Hence, our main goal was to
create a correct implementation of the general hierarchical linear
regression structure defined on SDT parameters. We argue that the
\code{bhsdtr} package for R \cite{rstatistical} provides exactly such
an implementation and we have made it publicly available at
\code{https://github.com/boryspaulewicz/bhsdtr}, together with the
annotated source code that was used to perform all the analyses and
produce all the figures presented in this paper.

In what follows, after introducing the most common version of the SDT
model, we describe its generalization, which can accommodate data from
rating experiments. Next, we explain briefly why, if a method of
inference for SDT models were to be of general use in psychology
studies, it is essential that it is based on a model equipped with the
general hierarchical linear regression structure. The \code{bhsdtr}
package meets this requirement thanks to a novel parametrization. We
describe this novel parametrization in great detail and explain how
reliance on relatively standard parametrization leads to problems in
the two other available implementations. We end the first part of this
paper with a formal definition of the model as implemented in
\code{bhsdtr}. The second part contains an overview of the package and
a tutorial in which we demonstrate how to use our method in
practice. Before we go any further, however, a note on terminology
seems in order.

In the context of hierarchical modelling, factors such as
participants, items, or replications are often referred to as
groups. In our opinion this naming convention may be confusing; a
single participant is both a group and a member of some group, while
at the same time the term "group" is perhaps most strongly associated
with study conditions, as in "experimental group". In this paper we
use the term "sampled factor" instead because, by virtue of being a
new term, it is unambiguous and seems descriptively correct: the term
"sampled factor" captures all the essential properties of such
variables, i.e., a nominal scale, the fact that values are sampled
from a larger population and are usually not of direct interest, as in
"this is only a sample", and that conclusions of statistical analysis
are meant to apply to the whole population of possible values.

@@

Ratings are accommodated by introducing additional criteria and
modelling a combined response $y$, which represents both the binary
classification decision and rating. The value of $y$ increases with
the strength of evidence in favor of the second stimulus class. For
example, if confidence is rated on a four-point scale, then $y = 1$
when a subject decides that a stimulus belongs to the first class with
certainty $4$, and $y = 5$ when a subject decides that a stimulus
belongs to the second class with certainty $1$. More formally, if $K$
is the number of possible combined responses, then a subject is
assumed to give response $k$ if $s \in (\bm{c}_{k-1},\bm{c}_k]$, where
$\bm{c}$ are the decision criteria, with $\bm{c}_0$ and $\bm{c}_K$
fixed at $-\infty$ and $+\infty$ respectively.

There is a good reason to collect the ratings and use the generalized
SDT model from Fig.~\ref{sdtr}, even when neither the ratings nor the
placement of criteria are relevant to the research problem. When
$K = 2$ (no ratings), the SDT model fits perfectly, regardless of
whether it is a reasonably good approximation to reality, because the
data and the model have the same dimensionality. This makes the
generalization to the case particularly important, as it is only when
$K > 2$ that the formal assumptions of the model (e.g., equal or
unequal variance) can be tested\footnote{In contrast to the formal
  assumptions, a psychological interpretation of the SDT parameters
  can be tested even when ratings are not available, e.g., by means of
  selective influence \cite{sternberg2001separate}}, which is often
done by comparing observed and predicted ROC curves.

When SDT models are used in psychology studies, researchers are
usually interested in relationships between SDT parameters and
additional measured or manipulated variables; a good example is the
dependence of $d'$ on stimulus strength. Also, in a great majority of
psychology studies in which classification tasks are used, the data
have a hierarchical structure, i.e., there are repeated measures for
each participant or item, and participants or items are only samples
from the target population. It seems worthwhile to explain why a
general-purpose method of inference for SDT models should accommodate
both kinds of situations.

@@

If data have a hierarchical structure, but variability due to
subjects, items, or other sampled factors is not accounted for,
estimates of average (fixed) effects are not guaranteed to be unbiased
and conclusions are not guaranteed to generalize to the target
population. The not uncommon practice of analyzing data aggregated
over sampled factors represents an extreme case of ignoring
hierarchical data structure. The invalidity of this approach in the
context of SDT was clearly illustrated with the results of
simulational studies by \citeA{morey2008problematic}; however,
strictly speaking, it requires such demonstrations only in order to
make a point about a specific case. Because SDT is a non-linear model,
by definition, when estimates of its parameters are based on data
aggregated over sampled factors, the expected values of these
estimates are not in general true population averages. Such estimates
are biased and inference about a target population based on them is
simply not valid. To give a concrete example, consider two unbiased
participants, one with $d'_1 = 2$ and one with $d'_2 = 4$. Their
expected average accuracy is given by
$(\Phi(d'_1/2) + \Phi(d'_2/2)) / 2 = .91$, which corresponds to
$d' = 2.68$, whereas their true average $d'$ is $3$.

@@

Another common source of problems is the separation of estimation of
non-linear model parameters from regression analyses which aim to
relate these parameters to measured or manipulated variables. When the
SDT parameters are estimated separately for each subject and
condition, and only later regressed on predictors of interest, a
number of additional issues may arise.

@@

The aforementioned problems can be dealt with by supplementing the SDT
model with the general hierarchical linear regression structure. For
reasons that we will now explain, this can only be achieved if the
model is substantially reparametrized.

@@

Both $d'$ and $c$ have the virtue of being directly interpretable in
terms of sensitivity and bias. However, both $d'$ and $c$ are
constrained: $d'$ is non-negative and, when there is more than one
criterion, the elements of the $\bm{c}$ vector are order restricted
($\bm{c}_{i+1} > \bm{c}_i$). Because normal distribution is unbounded,
the hierarchical linear regression structure in which random effects
are normally distributed and fixed effects are represented by
unconstrained parameters can only be defined on unconstrained
parameters. We provide examples of the problems that may otherwise
arise in the following summary of hierarchical SDT implementations.

We are aware of only two attempts at creating a general-purpose method
of inference for hierarchical SDT models with ratings. One is the
Gibbs sampler proposed by \citeA{morey2008problematic} and the other
is the Hierarchical Meta-d' model (HMeta-d') proposed by
\citeA{hmetad}. The HMeta-d' model is a hierarchical version of the
meta-d' model \cite{maniscalco2012signal}, which in turn is a
generalization of the SDT model that allows for a separate
"meta-sensitivity" to account for possible discrepancies between a
binary stimulus classification (referred to as a type 1 task) and the
associated rating task (referred to as a type 2 or meta-cognitive
task). We consider HMeta-d' here because it reduces to the SDT model
with ratings when the type 1 and type 2 sensitivities are equal.

The Gibbs sampler created by \citeA{morey2008problematic} allows for
at most two sampled factors to have independent normally distributed
random effects on the evidence distribution means. Unlike $d'$, each
evidence distribution mean considered in isolation is an unconstrained
parameter, but the mean of the second evidence distribution is by
definition greater than ($d' > 0$) or equal to ($d' = 0$) the mean of
the first. The authors explicitly admit that their algorithm does not
enforce this restriction because, as they claim in the paper, it would
greatly complicate analysis. At the same time the authors fail to
mention that an immediate consequence of allowing for negative $d'$
values is that the resulting posterior distribution loses its intended
interpretation since it can have a non-zero mass when $d' < 0$. The
outermost criteria are fixed at $0$ and $1$, and the ordering
restriction is enforced by assuming that the likelihood is $0$
whenever $\bm{c}_{i+1} \leq \bm{c}_i$. As the authors explain, because
a sampled factor can have independent random effects on the evidence
distribution means, it can have an effect on all the criteria:
shifting both means by the same amount in the same direction is
equivalent to keeping the sensitivity intact, while shifting the
criteria relative to the evidence distributions. However, the elements
of the criteria vector cannot be affected differently by the same
sampled factor, which is an unrealistic restriction.

In HMeta-d' the hierarchical structure is restricted to normally
distributed random intercepts of one sampled factor. Perhaps the most
problematic aspect of the HMeta-d' model, apart from the fact that the
$d'$ parameter is allowed to assume negative values in this model
also, is the representation of the criteria. Each element of the
criteria vector has an associated independent normal distribution,
which does allow for criteria random effects, but does not enforce the
necessary ordering restriction. The elements of this vector are sorted
to obtain another criteria vector, and it is this sorted criteria
vector that is used to model the conditional combined response
distributions. Consequently, the model does contain parameters
representing the actual, order-restricted criteria, but, because
sorting is not injective, the space of the actual criteria is only
loosely related (i.e., not isomorphic) to the space of the
unrestricted criteria vectors that are associated with the
hierarchical structure.

@@

It is often a good idea to multiply all the criteria by a value
greater than $1$, which is equivalent to making the mapping
distribution wider. This tends to even out values of $\bm{\gamma}$ by
preventing the outermost areas under the mapping distribution curve
from becoming very small, especially when the criteria are widely
spread, as can happen for moderate to large $d'$ values. This feature
is implemented in the \code{bhsdtr} package by introducing a scaling
factor.

@@

The \code{bhsdtr} package implements the model described in the
previous section in the Stan modelling language because it uses a
state-of-the-art adaptive Hamiltonian Monte Carlo sampling algorithm
which often handles high-dimensional correlated posteriors better than
a Gibbs sampler. Our package is essentially a collection of documented
functions: The \code{aggregate\_responses} function aggregates data as
much as possible for efficiency, but without distorting the
hierarchical structure. The \code{make\_stan\_model} function creates a
model definition in the Stan language. The Stan code produced by the
\code{make\_stan\_model} function can be fitted as is or modified by the
user if needed, e.g., to change the prior distributions or to drop the
equal variance assumption. The \code{make\_stan\_data} function creates
regression model matrices and other data structures required by the
model created using the \code{make\_stan\_model} function. Finally, the
\code{plot\_sdt\_fit} function can be used to visually assess the fit of
the model by creating publication-ready ROC curve plots or response
distribution plots with posterior predictive intervals calculated for
the chosen $\alpha$ level.

@@

To make full use of the bhsdtr package functionality, three
non-standard R packages are required, namely \code{rstan},
\code{plyr}, and \code{ggplot2}. We recommend using the
\code{devtools} package to install the \code{bhsdtr} package directly
from the github repository. This will automatically install any
missing required packages:

@@

The main purpose of the aggregation step is to improve the efficiency
of sampling from the posterior distribution. When data are aggregated
in this way, the likelihood for each condition $\times$ participant
combination has to be computed only once rather than as many times as
there are trials per condition per participant. Note that if there are
other sampled factors present in the data (e.g., items, replications,
etc.), then these factors also have to be specified at this stage to
preserve the hierarchical data structure. The
\code{aggregate\_responses} function creates a list with three
components. The \code{data} component is a data frame containing
additional preserved variables, the \code{stimulus} component is the
stimulus class variable, and the \code{counts} component is an
$N\times K$ matrix of combined response counts, where $N$ is the
number of data points and $K$ is the number of possible combined
response values.

>>>>

However, because the SDT model is non-linear, variability in its
parameters due to factors which are usually study-specific and of no
direct interest to the researcher such as participants or items has to
be accounted for, otherwise the estimates of SDT parameters are
biased. As we explain later in this paper, none of the available
methods of inference for hierarchical SDT models that we are aware of
addresses this problem correctly in it's generality. Hence, our main
goal was to create a correct implementation of the general
hierarchical linear regression structure defined on SDT parameters. We
argue that the \code{bhsdtr} package for R \cite{rstatistical}
provides exactly such an implementation and we have made it publicly
available at \code{https://github.com/boryspaulewicz/bhsdtr}, together
with the annotated source code that was used to perform all the
analyses and produce all the figures presented in this paper.

In what follows, after introducing the most common version of the SDT
model, we describe its generalization, which can accommodate data from
rating experiments. Next, we explain briefly why, if a method of
inference for SDT models were to be of general use in psychology
studies, it is essential that it is based on a model equipped with the
hierarchical linear regression structure. The \code{bhsdtr} package
meets this requirement thanks to a novel parametrization. We describe
this novel parametrization and explain how reliance on some popular
parametrizations leads to problems in the two other available
implementations. We end the first part of this paper with a formal
definition of the model as implemented in \code{bhsdtr}. The second
part contains an overview of the package and a tutorial in which we
demonstrate how to use our method in practice. Before we go any
further, however, a note on terminology seems in order.

In the context of hierarchical modelling, factors such as
participants, items, or replications are often referred to as
groups. In our opinion this naming convention may be confusing; a
single participant is both a group and a member of some group, while
at the same time the term "group" is perhaps most strongly associated
with study conditions, as in "experimental group". In this paper we
use the term "sampled factor" instead because, by virtue of being a
new term, it is unambiguous and seems descriptively correct: the term
"sampled factor" seems to capture essential properties of such
variables, i.e., a nominal scale, the fact that values are sampled
from a larger population and are usually not of direct interest, as in
"this is only a sample", and that conclusions of statistical analysis
are meant to apply to the whole population of possible values.

@@

Ratings are accommodated by introducing additional criteria and
modelling a combined response $y$, which represents both the binary
classification decision and rating as a single number. The value of
$y$ increases with the strength of evidence in favor of the second
stimulus class. For example, if confidence is rated on a four-point
scale, then $y = 1$ when a subject decides that a stimulus belongs to
the first class with certainty $4$, and $y = 5$ when a subject decides
that a stimulus belongs to the second class with certainty $1$. More
formally, if $K$ is the number of possible combined responses, then a
subject is assumed to give response $k$ if
$s \in (\bm{c}_{k-1},\bm{c}_k]$, where $\bm{c}$ are the decision
criteria, with $\bm{c}_0$ and $\bm{c}_K$ fixed at $-\infty$ and
$+\infty$ respectively.

There is a good reason to collect the ratings and use the generalized
SDT model from Fig.~\ref{sdtr}, even when neither the ratings nor the
placement of criteria are relevant to the research problem. When
$K = 2$ (no ratings), the SDT model fits perfectly, regardless of
whether it is a reasonably good approximation to reality, because the
data and the model have the same dimensionality. This makes the
generalization to the $K > 2$ case particularly important, as it is
only when $K > 2$ that the formal assumptions of the model (e.g.,
equal or unequal variance) can be tested\footnote{In contrast to the
  formal assumptions, a psychological interpretation of the SDT
  parameters can be tested even when ratings are not available, e.g.,
  by means of selective influence \cite{sternberg2001separate}}, which
is often done by comparing observed and predicted ROC curves.

When SDT models are used in psychology studies, researchers are
usually interested not in the values of the SDT parameters themselves,
but in relationships between these parameters and additional measured
or manipulated variables; a good example is the dependence of $d'$ on
stimulus strength. Also, in a great majority of psychology studies in
which classification tasks are used, the data have a hierarchical
structure, i.e., there are repeated measures for each participant or
item, and participants or items are only samples from the target
population. A general-purpose method of inference for SDT models
should accommodate both kinds of situations.

@@

If data have a hierarchical structure, but variability due to
subjects, items, or other sampled factors is not accounted for,
estimates of average (fixed) effects are not guaranteed to be unbiased
and conclusions are not guaranteed to generalize to the target
population. The not uncommon practice of analyzing data aggregated
over sampled factors represents an extreme case of ignoring
hierarchical data structure. The invalidity of this approach in the
context of SDT was clearly illustrated with the results of
simulational studies by \citeA{morey2008problematic}; however,
strictly speaking, it requires such demonstrations only in order to
illustrate a mathematical truth in a specific case. Because SDT is a
non-linear model, by definition, when estimates of its parameters are
based on data aggregated over sampled factors (e.g., $d'$ estimated
for hits and false alarms averaged over subjects), the expected values
of these estimates (e.g., what the calculated $d'$ actually estimates)
are not in general true population averages (e.g., true $d'$ in some
condition). Such estimates are biased and inference about a target
population based on them is simply not valid. To give a concrete
example, consider two unbiased participants, one with $d'_1 = 2$ and
one with $d'_2 = 4$. Their expected average accuracy is given by
$(\Phi(d'_1/2) + \Phi(d'_2/2)) / 2 = .91$, which corresponds to
$d' = 2.68$, whereas their true average $d'$ is $3$.

@@

Another common source of problems is the separation of estimation of
non-linear model parameters from regression analyses which aim to
relate these parameters to measured or manipulated variables. When the
SDT parameters are estimated separately for each subject and
condition, and only later are these estimates regressed on predictors
of interest, a number of additional issues may arise.

@@

The aforementioned problems can be dealt with by supplementing the SDT
model with the hierarchical linear regression structure. This can only
be achieved if the model is substantially reparametrized.

@@

Both $d'$ and $c$ have the virtue of being directly interpretable in
terms of sensitivity and bias. However, both $d'$ and $c$ are
constrained: $d'$ is non-negative and, when there is more than one
criterion, the elements of the $\bm{c}$ vector are order restricted
($\bm{c}_{i+1} > \bm{c}_i$). Hierarchical linear regression structure
can only be defined on unconstrained parameters, because random
effects are assumed to be normally distributed and normal distribution
is unbounded and because fixed effects are represented by
unconstrained parameters. We provide examples of the problems that may
arise when the SDT model is not correctly reparametrized in the
following summary of two hierarchical SDT implementations.

We are aware of two failed attempts at creating a general-purpose
method of inference for hierarchical SDT models with ratings. One is
the Gibbs sampler proposed by \citeA{morey2008problematic} and the
other is the Hierarchical Meta-d' model (HMeta-d') proposed by
\citeA{hmetad}. The HMeta-d' model is a hierarchical version of the
meta-d' model \cite{maniscalco2012signal}, which in turn is a
generalization of the SDT model that allows for a separate
"meta-sensitivity" to account for possible discrepancies between a
binary stimulus classification (referred to as a type 1 task) and the
associated rating task (referred to as a type 2 or meta-cognitive
task). We consider HMeta-d' here because it reduces to the SDT model
with ratings when the type 1 and type 2 sensitivities are equal.

The Gibbs sampler created by \citeA{morey2008problematic} allows for
at most two sampled factors to have independent normally distributed
random effects on the evidence distribution means. Unlike $d'$, each
evidence distribution mean considered in isolation is an unconstrained
parameter, but the mean of the second evidence distribution is by
definition greater than ($d' > 0$) or equal to ($d' = 0$) the mean of
the first. The authors explicitly admit that their algorithm does not
enforce this restriction because, as they claim in the paper, it would
greatly complicate analysis. At the same time the authors fail to
mention that an immediate consequence of allowing for negative $d'$
values is that the resulting posterior distribution loses its intended
interpretation since it can have a non-zero mass when $d' < 0$. The
outermost criteria are fixed at $0$ and $1$, and the ordering
restriction is enforced by assuming that the likelihood is $0$
whenever $\bm{c}_{i+1} \leq \bm{c}_i$. As the authors explain, because
a sampled factor can have independent random effects on the evidence
distribution means, it can have an effect on all the criteria:
shifting both means by the same amount in the same direction is
equivalent to keeping the sensitivity intact, while shifting the
criteria relative to the evidence distributions. However, the elements
of the criteria vector cannot be affected differently by the same
sampled factor, which is an unrealistic restriction. Participants
differ in how they place the criteria just as they differ in their
sensitivity and it is quite impossible for all the participants to
place the criteria in the same distance from each other.

In HMeta-d' the hierarchical structure is restricted to normally
distributed random intercepts of one sampled factor. In the HMeta-d'
model the $d'$ parameter is allowed to assume negative values also,
but the most problematic aspect of this implementation is again the
representation of the criteria. Each element of the criteria vector
has an associated independent normal distribution, which does allow
for criteria random effects, but does not enforce the necessary
ordering restriction. The elements of this vector are sorted to obtain
another criteria vector, and it is this sorted criteria vector that is
used to model the conditional combined response
distributions. Consequently, the model does contain parameters
representing the actual, order-restricted criteria, but, because
sorting is not injective, the space of the actual criteria is only
loosely related (i.e., not isomorphic) to the space of the
unrestricted criteria vectors that are associated with the
hierarchical structure.

@@

It is often a good idea to multiply all the criteria by a value
greater than $1$, which is equivalent to making the mapping
distribution wider. This tends to even out values of $\bm{\gamma}$ by
preventing the outermost areas under the mapping distribution curve
from becoming very small relative to ares delineated by adjacent pairs
of non-outermost criteria. This is especially important when the
criteria are widely spread, as can happen for moderate to large $d'$
values. This feature is implemented in the \code{bhsdtr} package by
introducing a criteria scaling factor.

@@

The \code{bhsdtr} package implements the model in the Stan modelling
language because it uses a state-of-the-art adaptive Hamiltonian Monte
Carlo sampling algorithm which often handles high-dimensional
correlated posteriors better than a Gibbs sampler. Our package is
essentially a collection of documented functions: The
\code{aggregate\_responses} function aggregates data as much as
possible for efficiency, but without distorting the hierarchical
structure. The \code{make\_stan\_model} function creates a model
definition in the Stan language. The Stan code produced by the
\code{make\_stan\_model} function can be fitted as is or modified by
the user if needed, e.g., to change the prior distributions or to drop
the equal variance assumption. The \code{make\_stan\_data} function
creates regression model matrices and other data structures required
by the model created using the \code{make\_stan\_model}
function. Finally, the \code{plot\_sdt\_fit} function can be used to
visually assess the fit of the model by creating publication-ready ROC
curve plots or response distribution plots with posterior predictive
intervals calculated for the chosen $\alpha$ level.

@@

To make full use of the \code{bhsdtr} package functionality, three
non-standard R packages are required, namely \code{rstan},
\code{plyr}, and \code{ggplot2}. We recommend using the
\code{devtools} package to install the \code{bhsdtr} package directly
from the github repository. This will automatically install the
package and any missing required packages:

@@

The main purpose of the aggregation step is to improve the efficiency
of sampling from the posterior distribution. When data are aggregated
in this way, the likelihood for each condition $\times$ participant
combination has to be computed only once rather than as many times as
there are trials per condition per participant. Note that if there are
other sampled factors present in the data (e.g., items, replications,
etc.) and the user decided to model the effects of these factors, then
these factors also have to be specified at this stage to preserve the
hierarchical data structure. The \code{aggregate\_responses} function
creates a list with three components. The \code{data} component is a
data frame containing additional preserved variables, the
\code{stimulus} component is the stimulus class variable, and the
\code{counts} component is an $N\times K$ matrix of combined response
counts, where $N$ is the number of data points and $K$ is the number
of possible combined response values.
